Time Series Analysis of Chaotic Dynamical Systems with Recurrent Neural Networks

### 2019-04-24 doc/paper/paper.tex
first draft complete

Neural Networks are known to be able to approximate any function given enough
data and time to train \cite{cybenko1989approximation}. In light of this, the
hope is that a deep learning based model, the ESN, will outperform ARIMA.

This is a paper about chaos.
% TODO: Intro
% TODO: mention \texttt{echonn}

@@

% TODO: Graph of converging LCE

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.6\textwidth}
        \includegraphics[width=\textwidth]{images/lorenz_nd_plot.png}
        \caption{Plot of each dimension of Lorenz System}
        \label{fig:lorenz_nd}
    \end{subfigure}
    \begin{subfigure}[b]{0.6\textwidth}
        \includegraphics[width=\textwidth]{images/lorenz_3d_plot.png}
        \caption{The Lorenz Butterfly}
        \label{fig:lorenz_3d}
    \end{subfigure}
    \caption{The Lorenz System Visuals}
    \label{fig:lorenz_butter}
\end{figure}

@@

The 3-body problem, or the generalized n-body problem, is a famous problem
which many have analyzed; most notably, Herni Poincaré wrote many papers
developing the area significantly \cite{chenciner2000remarkable}. One of
Poincaré's solutions, known as the circular restricted 3-body problem,
demonstrates chaotic properties \cite{oestreicher2007history}. Poincaré
performs several reductions of the original 3-body problem in order to come
to a much more useful form of the problem. The mass of one of the objects is
assumed to be negligible. This is the case for Sun-Earth-Moon orbits: the
moon has little effect on the orbit of the sun. Another aspect is the
rotating reference frame. Since the mass of one of the bodies is negligible,
the problem simplifies to first solving a two body problem. While much easier
to solver, Poincaré went further to stop the motion of the two planets by
using a rotating reference frame. Rotating with the planets, the problem can
be modeled seemingly by two static planets and a third mass-less planet
orbiting around those in a two dimensional plane.

@@

The N body problem is defined as follows \cite{chenciner2000remarkable}:

@@

@@

Although the notation in both papers was different, the equations were the
same. It's also worth mentioning that Eberle's paper was for his Masters
Thesis, and Frnka's paper has no details associated with it, but I was able
to find a mathematician with the exact name who achieved his doctorate. With
that said, I believe the math is reliable, and the simulations seem to work
and at least demonstrate chaotic properties (the only requirement for this
project). With my lacking physics knowledge, I was unable to derive these
equations myself from the assumptions made by Henri Poincaré outlined in
\cite{chenciner2000remarkable}.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=.5\linewidth]{doc/paper/images/lorenz/full_differential.png}
    \caption{A plot of the differences between time steps of an approximation of the Lorenz Butterfly}
    \label{fig:lorenz_diff}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=.5\linewidth]{doc/paper/images/doub_pend/full_differential.png}
    \caption{The entire difference dataset}
    \label{fig:doub_pend_diff}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=.5\linewidth]{doc/paper/images/r3body/full_differential.png}
    \caption{The entire difference dataset}
\end{figure}

@@

>>>>

Chaotic dynamical systems have been an important area of study
especially in fields of atmospheric sciences. However, mathematical
models are not always available, and if they are, very small errors in
measurements can amplify very quickly making the prediction useless.

Deep learning techniques are known to be able to approximate any
function given enough data and resources
\cite{cybenko1989approximation}. 
In light of this, the hope is that a deep learning based
model, the Echo State Network, will be able to capture the 
complexity of chaotic systems and forecast accurately.

Using a tool created by me to perform the analysis, \echonn, I will
explore this problem and apply the Echo State Network to three classic
chaotic problems. The python module can numerically solve these 
dynamical systems, provides an echo state network implemented by me,
and has classes to help with experimentation and visualization.

One of the first chaotic systems is
Henri Poincaré's solution to the restricted circular three body
problem, a very useful simplification of the popular three body
problem \cite{chenciner2015poincare}. The second system will be the
Lorenz equation, researched by the father of chaos, Edward Lorenz
\cite{lorenz1963deterministic}. 
The system is commonly known as the Lorenz Butterfly for its shape, and
it's the origin of the butterfly effect commonly used in pop culture.
The last dynamical system is the double pendulum. This system is very
commonly cited for its very random looking behavior despite being a
completely deterministic system.

Chaos of the systems will be measured using the systems' maximum
Lyapunov Characteristic Exponents, and data for the systems is
collected from numerical solutions of the dynamical system's governing
ODE equations and randomly generated initial conditions.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=.6\linewidth]{doc/paper/images/lorenz/lce_converge.png}
    \caption{Convergence of LCE with large $t$}
    \label{fig:lce_conv_lorenz}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{images/lorenz_nd_plot.png}
        \caption{Plot of each dimension of Lorenz System}
        \label{fig:lorenz_nd}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{images/lorenz_3d_plot.png}
        \caption{The Lorenz Butterfly}
        \label{fig:lorenz_3d}
    \end{subfigure}
    \caption{The Lorenz System Visuals}
    \label{fig:lorenz_butter}
\end{figure}

@@

The 3-body problem, or the generalized n-body problem, is a famous problem
which many have analyzed; most notably, Herni Poincaré wrote many papers
developing the area significantly \cite{chenciner2015poincare}. One of
Poincaré's solutions, known as the circular restricted 3-body problem,
demonstrates chaotic properties \cite{oestreicher2007history}. Poincaré
performs several reductions of the original 3-body problem in order to come
to a much more useful form of the problem. The mass of one of the objects is
assumed to be negligible. This is the case for Sun-Earth-Moon orbits: the
moon has little effect on the orbit of the sun. Another aspect is the
rotating reference frame. Since the mass of one of the bodies is negligible,
the problem simplifies to first solving a two body problem. While much easier
to solver, Poincaré went further to stop the motion of the two planets by
using a rotating reference frame. Rotating with the planets, the problem can
be modeled seemingly by two static planets and a third mass-less planet
orbiting around those in a two dimensional plane.

@@

The N body problem is defined as follows \cite{chenciner2015poincare}:

@@

An interesting solution (originally calculated by Carles Simó) can be
found where the trajectories of the bodies
are periodic and form a figure 8 \cite{chenciner2000remarkable}.

The gravitational constant, $G$ was redefined to equal 1, the masses of the 
objects were equal to 1, and the initial conditions are as follows:

\begin{align*}
    \rv_1 &= (0.97000436, -0.24308753) \\
    \rv_2 &= -\rv_1 \\
    \rv_3 &= (0, 0) \\
    \rv_1' &= -\rv_3 / 2 \\
    \rv_2' &= -\rv_3 / 2 \\
    \rv_3' &= (-0.93240737, -0.86473146) \\
\end{align*}

However, the resulting solution doesn't demonstrate chaotic properties. The LCE 
convergence chart (Figure \ref{fig:lce_fig_8}) shows the LCE converging to zero.

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/3body/fig8_short.png}
        \caption{Figure 8 solution to 3 body problem}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\linewidth]{doc/paper/images/3body/lce_converge.png}
        \caption{Nonchaotic LCE convergence}
        \label{fig:lce_fig_8}
    \end{subfigure}
    \caption{Nonchaotic Figure 8 Solution to 3 Body}
\end{figure}

@@

Although the notation in both papers was different, the equations were the
same. It's also worth mentioning that Eberle's paper was for his Masters
Thesis, and Frnka's paper has no details associated with it, but I was able
to find a mathematician with the exact name who achieved his doctorate. With
that said, I believe the math is reliable, and the simulations seem to work
and at least demonstrate chaotic properties (the only requirement for this
project). With my lacking physics knowledge, I was unable to derive these
equations myself from the assumptions made by Henri Poincaré outlined in
\cite{chenciner2015poincare}.

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\linewidth]{doc/paper/images/lorenz/full_differential.png}
        \caption{A plot of the differences between time steps}
        \label{fig:lorenz_diff}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/lce_converge.png}
        \caption{LCE converges to 0.935 for the Lorenz system}
    \end{subfigure}
    \caption{The difference butterfly and LCE convergence}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\linewidth]{doc/paper/images/doub_pend/full_differential.png}
        \caption{A plot of the differences between time steps}
        \label{fig:doub_pend_diff}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/lce_converge.png}
        \caption{LCE converges to 0.601 for the Double Pendulum system}
    \end{subfigure}
    \caption{The difference graph and LCE convergence}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\linewidth]{doc/paper/images/r3body/full_differential.png}
        \caption{A plot of the trajectory of third body}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/r3body/lce_converge.png}
        \caption{The LCE is 0.176 for the Restricted 3 Body Problem}
    \end{subfigure}
    \caption{The orbital pattern of the restricted 3 body problem}
\end{figure}

@@

The degree of chaotic behavior was measured by the largest Lyapunov
Characteristic Exponent of the system, and three dynamical systems were
used: Lorenz's convection equation, the Double Pendulum system,
and Henri Poincaré's solution to the Restricted Circular Three Body
Problem. The performance of the Echo State Network was determined from
RMSE of the trajectory after some time. The number of the Lyapunov
times that the model was able to predict accurately determined strength
of the ESN.

The Lorenz equation's trajectory was reported to have an LCE of 0.935,
and the predictability of the model was around 1 Lyapunov time for the
best performers. In general, the ESN forecasted poorly on the Lorenz
model due to the shape of the attractor. The ESN struggled to predict
accurately which "wing" to follow and usually diverged at the
intersection point.

The double pendulum performed the worst with an LCE of 0.601.
Its best predictors were
unable to predict at least one Lyapunov time. Much of the pendulums'
motion was small with occasionally very large spikes of motion. There
wasn't enough data for the ESN to capture the complexity of the signals
which trigger the large spikes of motion. When the ESN begins to
predict spikes (often incorrectly), the RMSE increases significantly.
It did however produce the correlation found between dimensions, so
some of the pendulum's patterns were captured successfully.

The restricted three body problem performed very well with a much
smaller LCE value of 0.176. Similar to the Lorenz, there was an
apparent pattern, but in this case, only a single point for the
trajectory to orbit around. Due to this, the performance of the ESN was
much better. Unbias testing RMSE would remain below 0.1 for several 
Lyapunov times, similar to the results of another paper with a small 
maximum LCE value.

It's possible that the ESN model needs more data to train on, or 
other methods need to be applied to improve the model's performance,
but it seems that highly chaotic systems can't be predicted very
easily, but lesser chaotic systems with attractor patterns can be
forecasted with more success.
